% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{longtable,booktabs,array}
\newcounter{none} % for unnumbered tables
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\section{PGI-RC Correction}

\subsection{What the correction does}

The gap \(\text{IORad} - \text{IOLib}\) equals the share of total
variance attributed to PGI EA. Because the PGI is a noisy proxy for the
true additive genetic factor \(g^*\), this share is attenuated. The
correction (Becker et al.~(2021)) rescales it upward using \(\rho > 1\),
the ratio of true genetic signal to observed PGI signal:

\[\rho = \sqrt{\frac{h^2_\text{SNP}}{R^2_\text{PGI}}} > 1\]

Because our decomposition distinguishes between-family and within-family
contributions, we need two separate correction factors ---
\(\rho_\text{between}\) and \(\rho_\text{within}\) --- each computed
from the corresponding \(h^2\) and \(R^2\) at that level.


\subsection{\texorpdfstring{Step 1. External heritability estimates
(\(h^2\))}{Step 1. External heritability estimates (h\^{}2)}}

We use two scenarios, differing in the source of \(h^2_\text{pop}\).
Both are for educational attainment measured in years of schooling.

\subsubsection{Scenario A --- Becker et al.~(2021),
WLS-specific}

\begin{itemize}
\tightlist
\item
  \(h^2_\text{pop} = 0.172\ (\text{SE} = 0.103)\): WLS-specific GREML
  estimate, Supplement Table 4.
\item
  \(h^2_\text{within}\): Becker does not report a within-family
  estimate. We impute it by scaling Howe's within/total ratio to
  Becker's larger \(h^2_\text{pop}\), on the assumption that the share
  of genetic variance transmitted within families is constant across
  populations:
\end{itemize}

\[h^2_{\text{within},\text{Becker}} = h^2_{\text{within},\text{Howe}} \times \frac{h^2_{\text{pop},\text{Becker}}}{h^2_{\text{pop},\text{Howe}}} = 0.04 \times \frac{0.172}{0.13} \approx 0.053\]

\begin{itemize}
\tightlist
\item
  \(h^2_\text{between} = h^2_\text{pop} - h^2_\text{within} \approx 0.119\)
\end{itemize}

\textbf{Modeling choice}: the proportional scaling assumption is not
testable from the available data. It produces a Becker scenario that is
internally consistent (within/total ratio matches Howe), but the true
within-family \(h^2\) for WLS could differ. Using Howe's
\(h^2_\text{within} = 0.04\) directly would overestimate
\(h^2_\text{between}\) (since Becker's total is larger), and was
therefore not used.

\subsubsection{Scenario B --- Howe et al.~(2022), Nature
Genetics}

The data sources included in the estimations include MoBa (see
Supplementary Table 1).

\begin{itemize}
\tightlist
\item
  \(h^2_\text{pop} = 0.13\ (95\%\ \text{CI: } 0.12\text{–}0.15)\)
\item
  \(h^2_\text{within} = 0.04\ (95\%\ \text{CI: } 0.02\text{–}0.05)\)
\item
  \(h^2_\text{between} = h^2_\text{pop} - h^2_\text{within} = 0.09\)
\end{itemize}

This scenario is internally consistent: both \(h^2\) components come
from the same study and population. It is the more conservative scenario
(smaller \(h^2\), smaller correction).

\subsubsection{Additive decomposition}

\(h^2_\text{pop}\) decomposes additively because \(h^2_\text{between}\)
and \(h^2_\text{within}\) are both fractions of total phenotypic
variance --- the numerators partition the total genetic variance with
the same denominator throughout:

\[h^2_\text{pop} = h^2_\text{between} + h^2_\text{within}\]


\subsection{\texorpdfstring{Step 2. Estimating \(R^2\) from the
data}{Step 2. Estimating R\^{}2 from the data}}\label{step-2.-estimating-r2-from-the-data}

We need two \(R^2\) estimates from the WLS sibling sample: one for the
between-family association between PGI and education, and one for the
within-family association.

\subsubsection{\texorpdfstring{Between-family
\(R^2\)}{Between-family R\^{}2}}\label{between-family-r2}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Residualize both education and PGI on
  \(\text{poly}(\text{birth\_year}, 2) \times \text{sex}\) and the top
  10 genetic principal components (PCs) (Becker et al.~2021).
\item
  Compute family means of both residuals.
\item
  Regress mean education residual on mean PGI residual using OLS.
  \(R^2_\text{between}\) is the \(R^2\) of this regression.
\end{enumerate}

\subsubsection{\texorpdfstring{Within-family
\(R^2\)}{Within-family R\^{}2}}\label{within-family-r2}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Residualize both education and PGI on
  \(\text{poly}(\text{birth\_year}, 2) \times \text{sex}\) only.
  \textbf{No PCs} are included.
\item
  Demean both residuals within each family, isolating within-family
  variation.
\item
  Regress demeaned education on demeaned PGI using OLS.
  \(R^2_\text{within}\) is the \(R^2\) of this regression.
\end{enumerate}

\subsubsection{ICC}

We estimate the intraclass correlation (ICC) from an intercept-only
mixed model, and use it in Step 3 to put \(h^2\) and \(R^2\) on the same
scale.

\subsubsection{Consistency check}

Because \(R^2_\text{between}\) and \(R^2_\text{within}\) use different
denominators (between- and within-family variance respectively), they
combine as a weighted average to give the population-level \(R^2\):

\[R^2_\text{pop} = \text{ICC} \times R^2_\text{between} + (1 - \text{ICC}) \times R^2_\text{within}\]

This differs from the additive \(h^2\) decomposition because \(R^2\)
components use level-specific denominators while \(h^2\) components all
use total variance. Becker et al.~report \(R^2_\text{pop} = 0.063\); our
estimate is \(R^2_\text{pop} = 0.065\).


\subsection{\texorpdfstring{Step 3. Computing
\(\rho\)}{Step 3. Computing \textbackslash rho}}\label{step-3.-computing-rho}

\(h^2_\text{between}\) and \(h^2_\text{within}\) from the literature are
fractions of \emph{total} phenotypic variance. \(R^2_\text{between}\)
and \(R^2_\text{within}\) from our data are fractions of \emph{between-}
and \emph{within-family} variance respectively. We convert \(R^2\) to
total-variance scale before dividing:

\[\rho_\text{between} = \sqrt{\frac{h^2_\text{between}}{\text{ICC} \times R^2_\text{between}}}, \qquad \rho_\text{within} = \sqrt{\frac{h^2_\text{within}}{(1 - \text{ICC}) \times R^2_\text{within}}}\]


\subsection{Step 4. Applying the
correction}

See \texttt{PGI\_RC\_derivations} for the full derivation.

\subsubsection{Notation}\label{notation}

{\def\LTcaptype{none} % do not increment counter
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.4211}}
  >{\raggedright\arraybackslash}p{(\linewidth - 2\tabcolsep) * \real{0.5789}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Symbol
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Definition
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\(\sigma^2(\tau_j')\) & Between-family variance from M2 (conditioning on
PGI + ascribed) \\
\(w\) &
\([\sigma^2(\delta_{ij}^*) - \sigma^2(\delta_{ij}')] / \sigma^2_\text{tot}\)
--- within-family share explained by ascribed characteristics \\
\(\Delta_\text{between}\) &
\([\sigma^2(\tau_j) - \sigma^2(\tau_j^*)] / \sigma^2_\text{tot}\) ---
observed between-family genetic share \\
\(\Delta_\text{within}\) &
\([\sigma^2(\delta_{ij}) - \sigma^2(\delta_{ij}^*)] / \sigma^2_\text{tot}\)
--- observed within-family genetic share \\
\(\rho_\text{between}\) &
\(\sqrt{h^2_\text{between} / (\text{ICC} \times R^2_\text{between})}\) \\
\(\rho_\text{within}\) &
\(\sqrt{h^2_\text{within} / ((1 - \text{ICC}) \times R^2_\text{within})}\) \\
\end{longtable}
}

\subsubsection{Corrected IOLib}\label{corrected-iolib}

\[\text{IOLib}_\text{RC} = \text{IOLib} - (\rho^2_\text{between} - 1)\,\Delta_\text{between}\]

Since \(\rho_\text{between} > 1\),
\(\text{IOLib}_\text{RC} < \text{IOLib}\): the noisy PGI left too much
variance in \(\sigma^2(\tau_j^*)\), over-attributing
shared-family-environment variance to the liberal measure.

\subsubsection{Corrected IORad}\label{corrected-iorad}

\[\text{IORad}_\text{RC} = \text{Sibcorr} + \rho^2_\text{within}\,\Delta_\text{within} + w\]

Since \(\rho_\text{within} > 1\),
\(\text{IORad}_\text{RC} > \text{IORad}\): the within-family genetic
contribution was underestimated.

\subsubsection{Verification}\label{verification}

\[\text{IORad}_\text{RC} - \text{IOLib}_\text{RC} = \rho^2_\text{between}\,\Delta_\text{between} + \rho^2_\text{within}\,\Delta_\text{within}\]

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{Methodological notes}\label{methodological-notes}

\textbf{Consider both \(h^2\) scenarios} The two sources (Becker
vs.~Howe) differ both in population and in method (GREML on WLS
vs.~within-family design in cohorts including MoBa). For WLS, Becker's
estimate is sample-specific but has a large SE (0.103) and requires an
assumption to obtain \(h^2_\text{within}\). Howe's estimate is more
precisely measured for the within-family component but comes from a
different population.

\textbf{Uncertainty in estimation.} In the current method the correction
factors \(\rho_{between}\) and \(\rho_{within}\) are held fixed at their
full-sample values across iterations. The variance components (and hence
Δ\_between, Δ\_within, w, Sibcorr) vary across bootstrap samples.
R²\_between and R²\_within are estimated from the full sample and held
fixed.

Considerations: First, bootstrap SEs capture sampling variability in the
variance components but not uncertainty in the external \(h^2\)
estimates. How to consider this too? Second, \(R^2\) could be recomputed
on each bootstrap resample. In practice the impact is small:
\(R^2_{between}\) is estimated from N\_families observations and is more
stable than the lmer variance components that drive most bootstrap
variance. In addition, having a single correction factor for reporting
is clearer conceptually.

\textbf{\(h^2_\text{within}\) scaling for Becker is an assumption.} The
proportional scaling
\(h^2_{\text{within},\text{Becker}} = 0.04 \times (0.172/0.13)\) assumes
the within/total heritability ratio is the same in WLS as in MoBa. The
alternative --- using Howe's \(h^2_\text{within} = 0.04\) unchanged with
Becker's \(h^2_\text{pop} = 0.172\) --- would be internally
inconsistent, artificially inflating \(h^2_\text{between}\).

\subsection{References}\label{references}

\begin{enumerate}
\def\labelenumi{\Alph{enumi})}
\tightlist
\item
  Becker, Joel, et al.~``Resource profile and user guide of the
  Polygenic Index Repository.'' Nature human behaviour 5.12 (2021):
  1744-1758.
\item
  Howe, Laurence J., et al.~``Within-sibship genome-wide association
  analyses decrease bias in estimates of direct genetic effects.''
  Nature genetics 54.5 (2022): 581-592.
\end{enumerate}

\end{document}
